Approximation as Differentiation

### 2025-07-02 main/first-order.tex
Lattice of list slices for listing example.

A programmer, or perhaps a language implementor, might want to embed additional ``approximation points'' into
composite data types, to allow for richer approximation structures. For example, for a list like $1 : 2 : []$
we might want to allow the partial lists $\bot$, $1 : \bot$ and $1 : 2 : \bot$ as approximants, in addition to
the approximants obtained by replacing elements with $\bot$. This can be accomplished by extending the syntax
with a \emph{lifting} type constructor $\tyLift$. Then, as an alternative to the usual definition of lists of
natural numbers as $\mu\alpha.1 + \synVar{nat} \times \alpha$, one could (within the language) define a type
of ``approximation lists'' $\mu\alpha.\tyLift\,(1 + \tyLift\,\synVar{nat} \times \alpha)$, with monadic
computations over approximated lists computing partial lists like $1 : \bot$. The lattice of approximants of
$1 : 2 : []$ now includes not only the ``shape-preserving'' approximations but also the prefixes obtained by
truncating successive tails with $\bot$:

  \begin{center}
    \begin{tikzpicture}
      \node (top) at (0,0) {$1 : 2 : []$};
      % row 2
      \node [below of=top] (ioi) {$1 : \bot : []$};
      \node [left of=ioi,xshift=-3cm] (oii) {$\bot : 2 : []$};
      \node [right of=ioi,xshift=3cm] (iio) {$1 : 2 : \bot$};
      % row 3
      \node [below of=ioi] (oio) {$\bot : 2 : \bot$};
      \node [left of=oio,xshift=-3cm] (ooi) {$\bot : \bot : []$};
      \node [right of=oio,xshift=3cm] (ioo) {$1 : \bot : \bot$};
      % row 4
      \node [below of=oio] (bot) {$[(\mathsf{a}, \bot), (\mathsf{b}, \bot), (\mathsf{a}, \bot)]$};

      % links
      \draw (top) -- (ioi);
      \draw (top) -- (oii);
      \draw (top) -- (iio);
      \draw (ioi) -- (oio);
      \draw (ioi) -- (ooi);
      \draw (ioi) -- (ioo);
    \end{tikzpicture}
  \end{center}

>>>>

A programmer, or perhaps a language implementor, might want to embed additional ``approximation points'' into
composite data types, to allow for richer approximation structures. For example, for a list like $1 : 2 : []$
we might want to allow the partial lists $\bot$, $1 : \bot$ and $1 : 2 : \bot$ as approximants, in addition to
the approximants obtained by replacing elements with $\bot$. This can be accomplished by extending the syntax
with a \emph{lifting} type constructor $\tyLift$. Then, as an alternative to the usual definition of lists of
natural numbers as $\mu\alpha.1 + \synVar{nat} \times \alpha$, one could (within the language) define a type
of ``approximation lists'' $\mu\alpha.\tyLift\,(1 + \tyLift\,\synVar{nat} \times \alpha)$, with monadic
computations over approximated lists computing partial lists like $1 : \bot$. The lattice of approximants of
$1 : 2 : []$ now includes not only all the ``shape-preserving'' approximations but also all prefixes obtained
by truncating successive tails with $\bot$:

  \begin{center}
    \tikzset{node distance=1cm}
    \begin{tikzpicture}
      \node (top) at (0,0) {$1 : 2 : []$};
      % row 2
      \node [below of=top] (ioi) {$1 : \bot : []$};
      \node [left of=ioi,xshift=-1.5cm] (oii) {$\bot : 2 : []$};
      \node [right of=ioi,xshift=1.5cm] (iio) {$1 : 2 : \bot$};
      % row 3
      \node [below of=ioi] (oio) {$\bot : 2 : \bot$};
      \node [left of=oio,xshift=-1.5cm] (ooi) {$\bot : \bot : []$};
      \node [right of=oio,xshift=1.5cm] (ioo) {$1 : \bot : \bot$};
      % row 4
      \node [below of=oio,xshift=-1cm] (ooo) {$\bot : \bot : \bot$};
      \node [below of=oio,xshift=1cm] (io) {$1 : \bot$};
      % row 5
      \node [below of=ooo,xshift=1cm] (oo) {$\bot : \bot$};
      % row 6
      \node [below of=oo] (o) {$\bot$};

      % links
      \draw (top) -- (ioi);
      \draw (top) -- (oii);
      \draw (top) -- (iio);
      \draw (ioi) -- (oio);
      \draw (ioi) -- (ooi);
      \draw (ioi) -- (ioo);
      \draw (oii) -- (oio);
      \draw (oii) -- (ooi);
      \draw (iio) -- (oio);
      \draw (iio) -- (ioo);
      \draw (ooi) -- (ooo);
      \draw (oio) -- (ooo);
      \draw (ioo) -- (ooo);
      \draw (ioo) -- (io);
      \draw (ooo) -- (oo);
      \draw (io) -- (oo);
      \draw (oo) -- (o);
    \end{tikzpicture}
  \end{center}
